Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-09-03 write-up/covid-writeup.tex
wrtieup taking form

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives many epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a genuine small simple random sample.  To address this, a procedure is presented that combines case-count data with small random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied this to the United States's only statewide random molecular sample collected from April 25--29th in Indiana.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

This paper has two main objectives.  The first objective is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.  The reason is straightforward: observed case-count data is plagued by selection bias and measurement error. Through simple calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.  In particular, the millions of tests in the US have a small effective sample size when compared to random sampling. These calculations demonstrate the importance of COVID-19 government-led data collection including random samples over time from the population.

@@

The second objective is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.  Random samples provide the necessary covariate information from a representative sample from the population to estimate selection propensities. These propensities are then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.  A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases (cite: book).  While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist and impact uncertainty quantification.

@@

After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} (cite: nonprob, Elliott) with random samples from the population.  (cite: Chen) considered combining nonprobability and probability samples.  Here, we also rely on  samples

A key issue is also that the random sampling will occur over time, potentially infrequently compared to the nonprobability sample.   Here
We then provide an extension of the statistical error decomposition

The coronavirus research literature has exploded. One large component of the volume of research is epidemiological forecasting.
These forecast models are often also used to estimate the effective reproduction number.  We extend (cite: Johndrow) to handle the more complex SIR dynamics and covariate information.

@@

In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.  We highlight how testing information is collected as well as demographic and comorbidity information.  In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.  We clarify the different between viral and antibody testing.  For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.

@@

\subsection{Probabilistic sample in Indiana}

@@

@@

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions. First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator. We extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data to try and avoid selection bias.  After adjusting for measurement-error, this leads to a model-based forecasts of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition. The method and statistical error decompositions guide our recommendations in Section XX.

@@

With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.  In causal inference~(cite: Robins), under certain assumptions, the observational data permits direct estimation of treatment propensity.  In the COVID-19 context, one does not observe those who decide not to be tested.  In order to estimate selection propensities, one needs auxiliary information.

To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.  Maximum likelihood estimation follows by maximizing
\begin{equation}
\label{eq:propensity}
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right) + \sum_{i=1}^N \log \left( 1 - \pi (X_i; \theta) \right).
\end{equation}
The first sum only involves individuals observed in the nonprobability sample.
The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.

@@

(cite: Chen)
Currently US testing does not collect additional covariate information; however, this need not be the case.  Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.  If this were the case, we could not longer rely on population-level summary statistics so an alternative is required.

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
$$
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
$$
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

@@

Let $\tilde I_j = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
$$
\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
$$
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.

@@

\subsubsection{Time-varying weights and random sampling}

Suppose we have two random samples at times $t$ and $t+k$ (i.e., $k$ days apart).

>>>>

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives many epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data with small random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count and death data as well as the United States's only statewide random molecular sample collected from April 25--29th.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

This paper has two main objectives.  The first objective is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.  The reason is straightforward: observed case-count data is plagued by selection bias and measurement error. Through simple calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.  In particular, the millions of tests in the US have a small effective sample size when compared to random sampling. These calculations demonstrate the importance of probabilistic sampling designs  over time for estimation of point prevalence and effective reproduction number.

@@

The second objective is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.  Random samples provide the necessary covariate information from a representative sample from the population to estimate selection propensities. These propensities are then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.  A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\cite{Hao2020,Song2020,Ray2020,Johndrow2020}.  While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist and impact uncertainty quantification.

@@

After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation. For any probability sampling design, the Horvitz-Thompson estimator~\citep{HT1952} incorporates design information via inverse-probability weights (IPW).  For nonprobability samples, the IPW estimator requires modelling the propensity scores.  Its use in the survey context is also referred to as the quasi-randomization approach~\citep{Elliott2017}. \cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.  \cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.  Here, we extend this approach to account for measurement-error as well as random sampling at multiple times. We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.

One core component of coronavirus research is epidemiological state-space modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements. Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.  For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.  Here, we combine the state-space SIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.  We generate epidemiological forecasts for point prevalence of each population strata.  We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.  A derived statistical error decomposition guides this discussion.

@@

In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.  We highlight how testing information is collected as well as demographic and comorbidity information.  In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.  We clarify the difference between viral and antibody testing.  For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.

@@

\subsection{Probabilistic samples in Indiana}

@@

{\bf Facebook symptoms. }
Finally, since XX, Facebook has conducted random sample testing ().  In particular, the COVID Symptom tracker provides an estimate of the number of individuals who have experienced any COVID-19 compatible symptoms.

@@

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions. First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator. We extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data to try and avoid selection bias.  After adjusting for measurement-error, this leads to a model-based forecasts of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition. The method and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~\citep{Elliott2017} techniques to adjust for selection bias.  In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.  In the COVID-19 context, one does not observe those who decide not to be tested.  In order to estimate selection propensities, one needs auxiliary information.

To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.  Maximum likelihood estimation follows by maximizing
\begin{equation}
\label{eq:propensity}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
\end{equation}
The first sum only involves individuals observed in the nonprobability sample.
The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.

@@

Currently US testing does not collect additional covariate information; however, this need not be the case.  Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.  If this were the case, we could not longer rely on population-level summary statistics so an alternative is required.

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
$$
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
$$
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population. Here, we only observe strata information.  Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

@@

Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
\begin{equation}
\label{eq:statdecomp2}
\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
$$
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.

Comparing~\eqref{eq:statdecomp} to~\eqref{eq:error} shows that weighting impacts the estimation error in two ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.

As demonstrated in the derivation of IPW approach, if the weights are correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then...

Thi
The weighting does not interact with the

@@

\subsubsection{Time-varying weights and infrequent random sampling}

Suppose we have two random samples at times $t$ and $t+k$ (i.e., $k$ days apart). Here, we extend the IPW approach by~\cite{Chen2019} to the temporal setting.

\begin{equation}
\label{eq:tvpropensity}
\sum_{t=1}^T \left[ \sum_{j=1}^N I_j \log \left( \frac{\pi (X_{j,t}; \theta)}{1-\pi(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_{j,t}; \theta) \right) \right].
\end{equation}

We could estimate separately if we had 1 random sample at each.  But perhaps it is better to smooth across; especially
